What We Eval in the Shadows

### 2021-09-03 paper/main.tex
Fix Section 6

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\renewcommand{\k}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother
\lstdefinelanguage{smalleR} {
  morekeywords={
    for,
    if,
    else,
    function
  },
  sensitive=true, % keywords are not case-sensitive
  morecomment=[l]{\#}, % l is for line comment
  morestring=[b]{"} % defines that strings are enclosed in double quotes
}
\lstset{
  language={smalleR},
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
%  framexleftmargin=1mm,
%  framexrightmargin=1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  alsoletter={., _, $},
  breaklines=true
}

@@

\citet{ecoop12} in their evaluation of the design of the R language, briefly
looked at the use of \eval in their corpus (1763 packages). They found 8500
\eval call sites and recorded 2M \eval calls. Because they included the base
libraries, their results were dominated by a few functions with
\c{match.arg}\footnote{A base function that matches an argument against a set of
candidate values, it uses \eval to get the candidate values.} being responsible
for over half of the recorded \eval calls. They discussed two use cases. The
first one is the evaluation of an expression extracted from a promise using the
\k{substitute} call in a new environment. The other is the invocation of a
function with dynamically computed names and arguments.

@@

\Eval accepts any value as an argument, but if the value is not an expression,
\eval returns it unchanged. Expressions account for \packageCodepercent of
arguments. More specifically, \packageSymbolpercent are symbols (variables such
as \k{x}), \packageLanguagepercent are language objects (expressions
representing function calls), and \packageExpressionpercent are expression
objects (lists of expressions). Further inspection reveals that \k{\_inherit}
accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
\k{ggplot2}.\footnote{This field is used to model inheritance in \k{ggproto},
  one of the many object-oriented systems in R.}

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute} and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected the top 117 sites which contribute to 90\% of the calls in the
corpus. We present examples from nine classes from this batch. The
classification is not exhaustive; we only study 117 call sites from the
\PkgEvalCallSites call sites in our corpus and many uses of \eval do not fit in
a specific category. The goal here is to demonstrate the wide variety of uses to
which \eval is put to use. This also speaks to the ubiquity of \eval in the R
ecosystem. The code snippets have been simplified for clarity.

\paragraph{Behavioral Extension} Sometimes it is convenient to allow users to
extend the behavior of a library function. \Eval allows clients to provide a
simple text string that shall be executed in the context of the function. The
example here allows users to provide formulas as strings, such as
\k{"exp(-x^2/2)"}, the contract being that the text can refer to some variable
\k{x} that will be set up by the called function.\footnote{The function is used
to generates a sequence of random numbers using the adaptive rejection sampling
algorithm and the string formula is used to describe the target density.} In
this case users could provide an anonymous function instead.

\begin{minipage}{.95\textwidth}
\medskip\underline{\c{AdapSamp::rARS}:}
\begin{lstlisting}
 function(n,formula) {
   p <- function(x) eval(parse(text=formula))
\end{lstlisting}\medskip
\end{minipage}

\paragraph{Code Generation}
\Eval can be used to generate boilerplate code that would either be tedious to
write by hand or that is only known after deployment. The example uses \eval to
automatically generate R bindings to C++ methods.

\begin{minipage}{.95\textwidth}
\medskip\underline{\c{Rcpp::.makeCppMethods}:}
\begin{lstlisting}
 function(methods,env) {
   for(what in cppMethods)
     methods[[what]] <-
       eval(substitute(function(...)CppObject$what(...)), env)
\end{lstlisting}\medskip
\end{minipage}

\paragraph{Extracting Varargs}
\Eval can be used to obtain the source code of the arguments passed to a vararg
parameter. One example is a function that iterates over the contents of the
vararg until it finds a non-null value. \k{substitute(alist(...))} yields an
expression that represents a call to \k{alist} but where the \k{...} are
expended with the arguments supplied to the function. \Eval then returns a
\k{list} with the unevaluated argument expressions.

\begin{minipage}{.95\textwidth}
\medskip\underline{\c{statnet.common::NVL}:}
\begin{lstlisting}
 function(...) {
   exprs <- eval(substitute(alist(...)))
\end{lstlisting}\medskip
\end{minipage}

\paragraph{Controlled Evaluation}
\Eval can be used to control when and if expressions are evaluated. For
example, \k{assertthat} provides users with the ability to make assertions and,
if they fail, produce legible error messages. The example here accepts a vararg
with a list of expressions. It extracts the associated code in the same way
as the extracting varags example and evaluates the expressions
sequentially in the caller's environment. If an expression does not hold, an
error is generated.

\begin{minipage}{.95\textwidth}
  \medskip\underline{\c {assertthat::see\_if}:}
\begin{lstlisting}
 function(..., env=parent.frame(), msg=NULL) {
   asserts <- eval(substitute(alist(...)))
   for (assertion in asserts) {
     res <- tryCatch({
       eval(assertion, env)
     }, assertError=function(e) { structure(FALSE, msg=e$message) })
\end{lstlisting}\medskip
\end{minipage}

\paragraph{Shadowing}
%%% TODO ?????????????//// <- left by Jan
Some packages use \eval to implement variations of the \k{base} R functions.
For example, the \k{igraph} package provides the \k{do_call} function shown
below. This is a variant of the builtin \k{do.call} function with different
semantics for resolving the function name.
%with the additional support for specifying the environment where the call
%should happen. -- wrong, do.call also allows specifying an envir
It constructs a call from a name and a list of arguments, and passes it to \eval.

\begin{minipage}{.95\textwidth}
  \medskip\underline{\c{igraph::do\_call}:}
\begin{lstlisting}
 function(f,...,args=list(),env=parent.frame()) {
   f <- substitute(f)
   eval(make_call(f,...,args), env)
 }
\end{lstlisting}\medskip
\end{minipage}

\paragraph{Boilerplate Removal} 
A variant of code generation is to let users write compact code that is expended
via \eval and \k{match.call}. The example uses \k{match.call} to access the
expression with which the function is called. The next three lines construct a
call to \k{stats::model.frame} with argument text of parameters \k{'formula'},
\k{'data'}, and \k{'weights'}. Then, the call expression is evaluated in the
parent environment. The use of \k{match.call} avoids repeated use of
\k{substitute} for extracting the argument expressions. It is extensible as new
parameter names can be easily added. This example illustrates one of the most
recurring uses of \eval by packages that do statistical modeling, and relates to
the \c{model.frame} shape in Table~\ref{tab:minimizedexpressions}.

\begin{minipage}{.95\textwidth}
  \medskip\underline{\c{survival::survfit.formula}:}
\begin{lstlisting}
 function(formula, data, weights, ...) {
   Call <- match.call()
   indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
   temp <- Call[c(1, indx)]
   temp[[1L]] <- quote(stats::model.frame)
   mf <- eval.parent(temp)
\end{lstlisting}\medskip
\end{minipage}

This combination of \k{match.call} and \eval for passing most of the current
call's argument to another function is also used to create wrappers. For
example, the \k{write.csv} uses it to forward most of the parameters to the
more general \k{write.table} function, only setting the CSV-related
parameters:

\begin{minipage}{.95\textwidth}
  \medskip\underline{\c{base::write.csv}:}
\begin{lstlisting}
 function (...)
 {
   Call <- match.call(expand.dots = TRUE)
   Call$sep <- ","
   Call$dec <- "."
   Call$qmethod <- "double"
   Call[[1L]] <- as.name("write.table")
   eval.parent(Call)
 }
\end{lstlisting}\medskip
\end{minipage}

The pattern eases maintenance of wrappers by avoiding the need to explicitly
list all the parameter names.\footnote{Private communication with a developer of
  R.} Wrapping functions in R is somewhat harder than in other languages because
of default arguments, varargs and missing arguments.

% % TODO: update to explain why the eval is needed or remove
% \paragraph{Formula Resolution}
% Many packages use \eval to evaluate formulas. The example extracts the
% \k{"formula"} attribute attached to the \k{pdMat} object and evaluates it.

% \begin{minipage}{.95\textwidth}
%   \medskip\underline{\c{nlme::formula.pdMat}:}
% \begin{lstlisting}
%  function(x, asList, ...) eval(attr(x, "formula"))
% \end{lstlisting}\medskip
% \end{minipage}

@@

\Eval allows functions to resolve symbols in different environments than the
current one. For example, the \k{data.table} package redefines the subsetting
operator\footnote{It defines it as a triplet \texttt{[i, j, k]}, where expression
  \texttt{j} is used to calculate a subset of rows from \texttt{i} grouped by \texttt{k}.}
for data frames to enable compact queries. The following  
computes the average departure delays on 2014 Delta flights for each
origin airport.

\begin{minipage}{.95\textwidth}
\medskip
\begin{lstlisting}
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
#    origin       V1
# 1:    LGA 11.35083
# 2:    EWR 14.96171
# 3:    JFK 12.45414
\end{lstlisting}\medskip
\end{minipage}

At its core, the \k{[} operator uses \k{substitute} to capture its operand
expressions, transforms them, and evaluates them using \eval in the given data
frame.\footnote{We omit the code as the implementation spans over 1700 lines,\\
  \indent \cf
\url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}} The
symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up
in the current environment, but in the \k{flights} data frame.

Non-standard scoping is an important use case of \eval in R. It enables the
design of functions which require less writing. Since R is used primarily
interactively, this trade-off is justifiable.

\paragraph{Domain-Specific Language}
A domain-specific language leverages R's grammar but redefines its semantics in
a suitable way. The example below illustrates string interpolation. The
\k{glue} function from the \k{glue} package extracts snippets of code enclosed
between braces and evaluates them using \eval, and finally splices their
results to construct the result.

\begin{minipage}{.95\textwidth}
\medskip
\begin{lstlisting}
> greeting <- "Hello"
> glue('{greeting} World!') 
# Hello World!
\end{lstlisting}\medskip
\end{minipage}

@@

\begin{minipage}{.95\textwidth}
\medskip
\begin{lstlisting}
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |> 
    glue_data("{origin} has average delay of {round(V1)} min")
# LGA has average delay of 11 min
# EWR has average delay of 15 min
# JFK has average delay of 12 min
\end{lstlisting}\medskip
\end{minipage}

\paragraph{Obfuscation}
A few packages use \eval to bypass the restrictions imposed by the \k{R CMD
  CHECK} tool. This tool enforces some well-formedness rules on a submitted
package before accepting it for inclusion in CRAN. Some static checks ensure
that the package code does not use certain restricted functions. Package authors
use \eval to obfuscate their code to get around this limitation. The example
below mutates the variable \k{s} in the package environment. However, that is
locked by default and \k{unlockBinding} is a restricted function. To work around
this, the call to unlock is done through \eval.

\begin{minipage}{.95\textwidth}
  \medskip\underline{\c{aibd::scalaEnsure}:}
\begin{lstlisting}
 function() {
   env <- parent.env(environment())
   eval(parse(text=paste0('unlockBinding("s", env)')))
   assign("s", s, envir=env)
   lockBinding("s", env)
\end{lstlisting}\medskip
\end{minipage}

>>>>

\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}   %% For author/year citations
\usepackage{listings,hyperref,multirow,paralist,xspace,url,wrapfig,tikz}
\usetikzlibrary{positioning,automata,fit,shapes.geometric,backgrounds,calc}
\usepackage{tabularx}
\input{macro}
\newcommand{\NOTE}[1]{{\it Note: #1}\xspace}
\newcommand{\authorcomment}[3]{\xspace\textcolor{#1}{{\bf #2} #3}\xspace}
\newcommand{\todo}[1]{\authorcomment{red}{TODO}{#1}}
\renewcommand{\k}[1]{\lstinline |#1|\xspace}
% cf. https://tex.stackexchange.com/a/144640
\makeatletter\let\expandableinput\@@input\makeatother
\lstdefinelanguage{smalleR} {
  keywords={
    for,
    if,
    else,
    function
  },
  sensitive=true, % keywords are case-sensitive
  morecomment=[l]{\#}, % l is for line comment
  morestring=[b]{"} % defines that strings are enclosed in double quotes
}
\lstset{
  language={smalleR},
  columns=flexible,
  captionpos=b,
  frame=single,
  framerule=0pt,
%  framexleftmargin=1mm,
%  framexrightmargin=1mm,
  tabsize=2,
  belowskip=0pt,
  basicstyle=\small\ttfamily,
  backgroundcolor=\color{LightGray},
  emphstyle=\sffamily,
  keywordstyle=\bfseries,
  commentstyle=\color{Gray}\em,
  stringstyle=\color{Gray},
  showspaces=false,
  showstringspaces=false,
  deletekeywords={see_if, $, _, .},
  % alsoletter={., _, $},
  %emph={eval.parent, evalq, eval, local},
  %emphstyle=\underbar,
  breaklines=true
}

@@

\citet{ecoop12} in their evaluation of the design of the R language, briefly
looked at the use of \eval in their corpus (1763 packages). They found 8500
\eval call sites and recorded 2M \eval calls. Because they included the base
libraries, their results were dominated by a few functions with
\k{match.arg}\footnote{A base function that matches an argument against a set of
candidate values, it uses \eval to get the candidate values.} being responsible
for over half of the recorded \eval calls. They discussed two use cases. The
first one is the evaluation of an expression extracted from a promise using the
\k{substitute} call in a new environment. The other is the invocation of a
function with dynamically computed names and arguments.

@@

\Eval accepts any value as an argument, but if the value is not an expression,
\eval returns it unchanged. Expressions account for \packageCodepercent of
arguments. More specifically, \packageSymbolpercent are symbols (variables such
as \k{x}), \packageLanguagepercent are language objects (expressions
representing function calls), and \packageExpressionpercent are expression
objects (lists of expressions). Further inspection reveals that \k{\_inherit}
accounts for \packageGgplotsymbolpercent of symbols and comes from one site in
\k{ggplot2}.\footnote{This field is used to model inheritance in \k{ggproto}, an
  object-oriented system used by \k{ggplot2}.}

@@

The R language was intended to be extensible. The combination of lazy
evaluation, \k{substitute}, and \k{eval} are some of the tools given to
developers to this end. To understand \emph{why} developers use \eval, we
manually inspected the top 117 sites, which contribute to 90\% of the calls in
the corpus. We present examples from nine classes from this batch. The
classification is not exhaustive; we present results from only 117 of the total
\PkgEvalCallSites call sites in our corpus, and many uses of \eval do not fit in
a specific category. The \eval usage patterns are described below.

\paragraph{Behavioral Extension} Sometimes it is convenient to allow users to
extend the behavior of a library function. \Eval allows clients to provide a
simple text string that is executed in the context of the function. The
example here allows users to provide formulas as strings, such as
\k{"exp(-x^2/2)"}, the contract being that the text can refer to some variable
\k{x} that will be set up by the called function.\footnote{The function is used
to generate a sequence of random numbers using the adaptive rejection sampling
algorithm and the string formula is used to describe the target density.}

\begin{lstlisting}[caption={\c{AdapSamp::rARS}}, captionpos=b]
 function(n,formula) {
   p <- function(x) eval(parse(text=formula))
\end{lstlisting}

\paragraph{Code Generation}
\Eval can be used to generate boilerplate code that would either be tedious to
write by hand or that is only known after deployment. The following code snippet
for the \k{Rcpp} package uses \eval to automatically generate R bindings to C++
methods.

\begin{lstlisting}[caption={\k{Rcpp::.makeCppMethods}}, captionpos=b]
 function(methods,env) {
   for(what in cppMethods)
     methods[[what]] <-
       eval(substitute(function(...)CppObject$what(...)), env)
\end{lstlisting}

\paragraph{Extracting Varargs}
\Eval is used to obtain the source code of the arguments passed to a vararg
(\k{...}) parameter. In the following example, \k{substitute(alist(...))} yields
an expression representing a call to \k{alist} with the \k{...} expanded to the
arguments supplied to the function. \Eval evaluates this expression and returns
a \k{list} of the unevaluated argument expressions.

\begin{lstlisting}[caption={\k{statnet.common::NVM}}, captionpos=b]
 function(...) {
   exprs <- eval(substitute(alist(...)))
\end{lstlisting}\medskip

\paragraph{Controlled Evaluation}
\Eval can be used to control when and if expressions are evaluated. For example,
\k{assertthat} provides users with the ability to make assertions and, if they
fail, produce legible error messages. The \k{assertthat::see_if} function
accepts a vararg with a list of expressions. It extracts these expressions
(\emph{Extracting Varargs} pattern) and evaluates them sequentially in the
caller's environment using \k{eval}. If an expression does not hold, an error is
generated.

\begin{lstlisting}[caption={\k{assertthat::see\_if}}, captionpos=b]
 function(..., env=parent.frame(), msg=NULL) {
   asserts <- eval(substitute(alist(...)))
   for (assertion in asserts) {
     res <- tryCatch({
       eval(assertion, env)
     }, assertError=function(e) { structure(FALSE, msg=e$message) })
\end{lstlisting}\medskip

\paragraph{Shadowing}
Some packages use \eval to implement variations of the \k{base} R functions. For
example, the \k{igraph} package provides the \k{do_call} function shown below.
This is a variant of the builtin \k{do.call} function with different semantics
for resolving the function name. It constructs a call from a name and a list of
arguments and passes it to \eval.

\begin{lstlisting}[caption={\k{igraph::do\_call}}, captionpos=b]
 function(f,...,args=list(),env=parent.frame()) {
   f <- substitute(f)
   eval(make_call(f,...,args), env)
 }
\end{lstlisting}\medskip

\paragraph{Boilerplate Removal} 
A variant of code generation is to let users write compact code that is expended
via \eval and \k{match.call}. The example uses \k{match.call} to access the
expression with which the function is called. The following three lines construct a
call to \k{stats::model.frame} with argument text of parameters \k{'formula'},
\k{'data'}, and \k{'weights'}. Then, the call expression is evaluated in the
parent environment. The use of \k{match.call} avoids repeated use of
\k{substitute} for extracting the argument expressions. It is extensible as new
parameter names can be easily added. This example illustrates one of the most
recurring uses of \eval by packages that do statistical modeling, and relates to
the \k{model.frame} shape in Table~\ref{tab:minimizedexpressions}.

\begin{lstlisting}[caption={\k{survival::survfit.formula}}, captionpos=b]
 function(formula, data, weights, ...) {
   Call <- match.call()
   indx <- match(c('formula', 'data', 'weights'), names(Call), nomatch=0)
   temp <- Call[c(1, indx)]
   temp[[1L]] <- quote(stats::model.frame)
   mf <- eval.parent(temp)
\end{lstlisting}\medskip

Wrapping functions in R is somewhat harder than in other languages because of
default arguments, varargs, and missing arguments. The combination of
\k{match.call} and \eval can be used to forward the current call's arguments to
another function without listing all the argument names. This pattern eases the
maintenance of wrappers. For example, the \k{base::write.csv} function passes
most of its arguments unchanged to the more general \k{base::write.table}
function.

\begin{lstlisting}[caption={\k{base::write.csv}}, captionpos=b]
 function (...) {
   Call <- match.call(expand.dots = TRUE)
   Call$sep <- ","
   Call$dec <- "."
   Call$qmethod <- "double"
   Call[[1L]] <- as.name("write.table")
   eval.parent(Call)
 }
\end{lstlisting}\medskip

@@

\Eval allows functions to resolve symbols in different environments than the
current one. For example, the \k{data.table} package redefines the subsetting
operator, \k{[}\footnote{Defined as a triplet \texttt{[i, j, k]}, where
  expression \texttt{j} is used to calculate a subset of rows from \texttt{i}
  grouped by \texttt{k}.}, to enable compact queries on data frames. The example
below shows the use of this operator on the \k{flights} data frame to compute
the average departure delays of Delta flights in 2014 by the origin airport. The
symbols \k{carrier}, \k{year}, \k{dep_delay} and \k{origin} are not looked up in
the current environment, but in the \k{flights} data frame.

\begin{lstlisting}[caption={\k{data.table::[}}, captionpos=b]
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin]
#    origin       V1
# 1:    LGA 11.35083
# 2:    EWR 14.96171
# 3:    JFK 12.45414
\end{lstlisting}\medskip

At its core, the \k{[} operator uses \k{substitute} to capture its operand
expressions, transforms them, and evaluates them using \eval in the given data
frame. We omit the implementation as it spans over 1700 lines at the time of
this writing\footnote{Its implementation can be found at:
  \url{https://github.com/Rdatatable/data.table/blob/master/R/data.table.R}}.

Non-standard scoping is an important use case of \eval in R. It enables the
design of functions that require less typing. Since R is used primarily
interactively, this trade-off is justifiable.

\paragraph{Domain-Specific Language}
A domain-specific language leverages R's grammar but redefines its semantics in
a suitable way. The example below illustrates string interpolation. The \k{glue}
function from the \k{glue} package extracts snippets of code enclosed between
braces, evaluates them using \eval, and splices their results to construct the
result.

\begin{lstlisting}[caption={\k{glue::glue}}, captionpos=b]
> greeting <- "Hello"
> glue('{greeting} World!') 
# Hello World!
\end{lstlisting}\medskip

@@

\begin{lstlisting}[caption={\k{glue::glue_data}}, captionpos=b]
> flights[carrier=="DL" & year == 2014, mean(dep_delay), by=origin] |> 
    glue_data("{origin} has average delay of {round(V1)} min")
# LGA has average delay of 11 min
# EWR has average delay of 15 min
# JFK has average delay of 12 min
\end{lstlisting}\medskip

\paragraph{Obfuscation}
A few packages use \eval to bypass the restrictions imposed by the \k{R CMD
  CHECK} tool. This tool enforces some well-formedness rules on a submitted
package before accepting it for inclusion in CRAN. Some static checks ensure
that the package code does not use certain restricted functions. Package authors
use \eval to obfuscate their code to get around this limitation. The example
below mutates the variable \k{s} in the package environment. However, that is
locked by default and unlocking it requires \k{unlockBinding} which is a
restricted function. To work around this, the call to \k{unlockBinding} is done
through \eval.

\begin{lstlisting}[caption={\k{aibd::scalaEnsure}}, captionpos=b]
 function() {
   env <- parent.env(environment())
   eval(parse(text=paste0('unlockBinding("s", env)')))
   assign("s", s, envir=env)
   lockBinding("s", env)
\end{lstlisting}\medskip
